Understanding the problem of when to scale

### 2024-04-19 presentations/understanding the problem.qmd
update presentation

::: columns
::: {.column width="50%"}

- Same data gave different indicator values.
- Reference values are at county scales (and probably only make sense at that scale)
- IBECA normalised the variables (with truncation) at plot scale
- NI normalised the variable (probably) at a regional scale after first taking a mean
:::

::: {.column width="50%"}

![](img/billberry.PNG){width=100%}

:::

:::

---

@@

---

@@

## It matters 
if we  aggregate (takes means of) of the scaled or unscaled variabel:

@@

- Positive (and negative) outliers in on area can not compensate for poor (or good) condition elsewhere
- Gives higher-resolution indicator values

- The reference value was maybe intended for that scale
- The indicator uncertainty becomes (apparently) smaller because the mean or sum for a region is less variable.
- Maybe, to many truncation events leds to accumulated dissplacement errors?

- The reference value is at the same scale as the variable
- The reference value is the same all over
- We scale locally before aggregating, and then truncate later
- We also want to aggregate the original variable further, e.g. to national level - will the indicator and variable remain correlated

## Why do we re-scale
*notes from Bård*

Rescaling is useful for making indeces, but why use it for individual indicators? [answer: it depends if we want to aggregate 'condition estimates' of actuall measurements]

## Spatially aggregating 
condition estimates (bottom) or actuall measurements (top)

@@

## Early scaling 
leads to cummutativity

@@

## Early scaling 
leads to cummutativity

@@

## Commutativty

## Alternativey,
we can aggregate the actual measurements. When would this make sense to do?

@@

## What about

- Scaling, aggregating, and then truncating
- Non-linear re-scaling functions with asymptotes

>>>>

::: columns
::: {.column width="50%"}
-   Same data gave different indicator values.
-   Reference values are at county scales (and probably only make sense at that scale)
-   IBECA normalised the variables (with truncation) at plot scale
-   NI normalised the variable (probably) at a regional scale after first taking a mean
:::

::: {.column width="50%"}
![](img/billberry.PNG){width="80%"}
:::
:::

@@

@@

## It matters

if we aggregate (takes means of) of the scaled or unscaled variabel:

@@

## Pathways

```{r}
source("../R/pathways.R")
pathways
```

## Glossery

**Scaling** = Linear range scaling, usually reducing the range of the variable by using one or two reference value

**Truncating** = Cutting of value above or below a given threshold. In our case, making the metrices bound between 0 and 1. Truncation is also a type of scaling.

**Re-scaling** = not used, but often used as a synonym to normalisation.

**Normalising** = a combination of scaling, followed by truncating (if needed). This lead to a non-linear transformation of the original variable and returns indicators that share the same scale.

**Non-linear scaling**: Scaling functions like truncation, sigmoid, exponential or break-point types.

## Spatially aggregating

condition estimates (bottom) or actual measurements (top)

@@

------------------------------------------------------------------------
## Commutativty

## Early scaling

leads to cummutativity

@@

## Early scaling

leads to cummutativity

@@

## Alternativey,

we can aggregate the actual measurements. When would this make sense to do?

@@

## Why do we normalise

*notes from Bård*

Rescaling is useful for making indeces, but why use it for individual indicators? \[answer: it depends if we want to aggregate 'condition estimates' of actuall measurements\]

-   High or low *outliers* can not compensate for low or high values (i.e. condition) elsewhere, respectively.
-   Gives higher-resolution indicator values and preserves the original spatial resolution of the variable.
-   Means that subsequent aggregation is carrying information about the actual condition, and not the raw variable.

-   The reference value was maybe intended for that scale
-   The indicator uncertainty becomes (apparently) smaller because the mean or sum for a region is less variable.
-   Maybe, too many truncation events leads to accumulated displacement errors?

-   The reference value is at the same scale as the variable
-   The reference value is the same all over
-   We also want to aggregate the original variable further, e.g. to national level - will the indicator and variable remain correlated
-   We want to use non-linear re-scaling functions with asymptotes
